---
title: "Example 5 : Data Visualization with Preprocessed Data"
output: html_document
---

# Overview

This example demonstrates how to create several types of data visualizations using the ggplot2 library in R based on data that has been preprocessed using CAS actions. The data used for this example is 28,347 crime incidents from 2021 provided by the City of Washington, DC.

The data can be obtained from https://support.sas.com/documentation/onlinedoc/viya/exampledatasets/Crime_Incidents_in_2021.csv and is originally from https://opendata.dc.gov/datasets/DCGIS::crime-incidents-in-2021/about.

# Load the SWAT Library and Connect to the CAS Server

Load the SWAT library and then create a connection to the CAS server using the CAS function and assign the CAS connection object to the object s. The first argument specifies the host name, and the second argument specifies the port. The host and port information provided below is an example.

```{r}
library(swat)
# change the host and port to match your site
s <- CAS("example.sas.com", 10065)
```

# Import R Packages

Import the tidyverse and lubridate packages in R.

```{r include=FALSE}
library(tidyverse)
library(lubridate)
```

# Load the Data

## Load the Data from a Caslib

The default method of loading data is to load the data from the data source portion of a caslib, which is known as a server-side load. This requires the data file to be saved in the active caslib (Casuser). Once the file has been saved to the caslib, use the table.loadTable action to load the Crime_Incidents_in_2021.csv CSV file from the data source portion of the caslib into memory as a CAS table named crimes.

```{r}
cas.table.loadTable(s, 
                    path="Crime_Incidents_in_2021.csv",
                    caslib="casuser",
                    casOut=list(name="crimes", 
                                replace=TRUE),
                    importOptions=list(fileType="CSV",
                                       encoding="latin1",
                                       guessrows=30000,
                                       vars=list(ward=list(type="VARCHAR"))))
```

## Load a Client-Side Data File into CAS

Another method of loading data into CAS memory is to load the data from an external source that is accessible to the CAS server. This example uses the cas.upload.file function to perform a client-side load.

```{r include=FALSE}
cas.upload.file(s,                                                                        
                "https://support.sas.com/documentation/onlinedoc/viya/exampledatasets/Crime_Incidents_in_2021.csv", 
                casOut=list(name="crimes", 
                            replace=TRUE),
                importOptions=list(fileType="CSV",
                                   encoding="latin1",
                                   guessrows=30000))
```

# Explore the Data

## Create a Reference to the In-Memory Table

The defCasTable function creates an R object that references the in-memory CASTable. This allows you to use functions with an R syntax on the in-memory table without needing to explicitly write the CAS action and also allows you to use this object in CAS actions without having to specify parameters such as the connection object, in-memory CASTable name, or caslib.

```{r}
tbl <- defCasTable(s, "crimes", caslib="casuser")
```

## Examine the Rows

Run the table.fetch action to retrieve the first five rows from the crimes table.

```{r}
cas.table.fetch(tbl,
                to=5)
```

## Examine the Columns

Run the table.columnInfo action to obtain metadata about the table. The result includes the names of columns, and information about each column, including its label (if applicable), type, length, and format. The results show that the columns are the appropriate types.

```{r}
cas.table.columnInfo(tbl)
```

## Examine Unique and Missing Values

Run the simple.distinct action to identify the number of distinct values and the number of missing values for each column.

```{r}
cas.simple.distinct(tbl)
```

# Visualize the Data

The ggplot function from the tidyverse set of packages can be used to visualize data that has been preprocessed using CAS actions. This example demonstrates how to create several common types of data visualizations, including a bar chart, histogram, box plot, and line plot.

## Create a Bar Chart

Create a bar chart that shows the number of crime incidents by the shift variable.

### Aggregate the Data

First, run the  simple.freq action to count the number of rows by shift and specify the $Frequency key to save the output table to a casDataFrame object named casdf_shift.

```{r}
casdf_shift <- cas.simple.freq(tbl,
                inputs=list("SHIFT"))$Frequency
```

### Convert the CAS Table to an R Data Frame

Convert the casDataFrame casdf_shift to an R DataFrame named df_shift. Use head to view rows from df_shift.

```{r}
df_shift <- data.frame(tbl_shift)
head(df_shift)
```

### Plot the Data

Use the ggplot function with the geom_col function to create a bar chart. Specify the R DataFrame df_shift as the first argument. Specify CharVar (shift) as the X-axis variable and Frequency as the Y-axis variable, and order the bars in descending order by the frequency variable.

```{r}
ggplot(df_shift, aes(x=reorder(CharVar, -Frequency), y=Frequency)) +
  geom_col(fill="lightblue") +
  labs(x = "Shift", 
       y = "Number of Crime Incidents", 
       title = "Number of Crime Incidents by Shift",
       subtitle = "Washington DC, 2021") +
  theme_classic()
```

## Create a Histogram

Create a histogram to show the distribution of crime incidents by the count of voting precincts.

### Aggregate the Data

Run the simple.freq action to count the number of rows by the voting precinct and specify the $Frequency key save the resulting output table as a casDataFrame object named casdf_vp.

```{r}
casdf_vp <- cas.simple.freq(s, 
                            table=list(caslib="casuser",
                            name="crimes"),
                            inputs=list("VOTING_PRECINCT"))$Frequency
```

### Convert the CAS Table to an R Data Frame

Convert the casDataFrame casdf_vp to an R DataFrame named df_vp. Use head to view rows from the DataFrame.

```{r}
df_vp <- data.frame(casdf_vp)
head(df_vp)
```

### Plot the Data

Use the ggplot function with the geom_histogram function to create a histogram to show the distribution of crime incidents. Specify the R DataFrame df_vp as the first argument and Frequency as the X-axis variable.

```{r}
ggplot(df_vp, aes(x=Frequency)) +
  geom_histogram(binwidth=50, color="black", fill="white") +
  ylim(0, 40) +
  labs(x='Number of Crime Incidents',
       y='Number of Voting Precincts',
       title='Distribution of the Frequency of Crime Incidents by Voting Precincts',
       subtitle='Washington DC, 2021')
```

## Create a Box Plot

Create a box plot that shows summary statistics for the number of crime incidents among neighborhood clusters by ward.

### Aggregate the Data

Load the aggregation CAS action set. Run the aggregation.aggregate action to count the number of crime incidents by neighborhood cluster and ward and save the resulting output table as CrimesByNCWard. Create a CAS table object named tbl_nc_ward to reference the CrimesByNCWard table.

```{r}
cas.builtins.loadActionSet(s, actionSet="aggregation")

cas.aggregation.aggregate(s,                                                     
                          table=list(name="crimes",
                                     caslib="casuser",
                                     groupby=list(list(name="NEIGHBORHOOD_CLUSTER"), list(name="WARD"))),                     
                          varSpecs=list(list(name="CCN",                          
                                             subset=list("N"))),
                          casOut=list(name="CrimesByNCWard",                        
                                      caslib='casuser',
                                      replace=TRUE))

tbl_nc_ward <- defCasTable(s, "CrimesByNCWard", caslib="casuser")
```

### Convert the CAS Table to an R Data Frame

Convert the CrimesByNCWard table to a casDataFrame using the CAS table object tbl_nc_ward, and then convert the casDataFrame casdf_nc_ward to an R DataFrame named df_nc_ward. This is necessary because the ggplot function cannot use a reference to a CAS table since a CAS table is on the server and not the client-side. Use head to view the rows. 

```{r}
casdf_nc_ward <- to.casDataFrame(tbl_nc_ward, obs=nrow(tbl_nc_ward))
df_nc_ward <- data.frame(casdf_nc_ward)
head(df_nc_ward)
```

### Plot the Data

Use the as.character function to convert the Ward variable to character type. Use ggplot to with the geom_boxplot function to create a box plot to show the summary statistics for number of crime incidents by Ward. Specify ward as the X-axis variable and X_CCN_Summary_NObs_ (number of rows) as the Y-axis variable.

```{r}
df_nc_ward$WARD <- as.character(df_nc_ward$WARD)
ggplot(data=df_nc_ward, aes(x=WARD, y=X_CCN_Summary_NObs_)) +
  geom_boxplot(fill="lightblue") +
  labs(x = 'Ward', 
       y = 'Number of Crime Incidents',
       title = 'Box Plot of Crime Incidents by Ward',
       subtitle = 'Neighborhood Clusters in Washington DC, 2021')
```

## Create a Line Plot

Create a line plot to show the number of crime incidents by month in 2021.

### Convert the CAS Table to an R Data Frame

Convert the crimes in-memory table to a casDataFrame using the CAS table object tbl, and then convert the casDataFrame casdf_crimes to an R DataFrame named df_crimes. Use head to view the rows.

```{r}
casdf_crimes <- to.casDataFrame(tbl, obs=nrow(tbl))
df_crimes <- data.frame(casdf_crimes)
head(df_crimes)
```

### Create a Date Variable

Use the mutate and as.Date functions to create a new data variable containing only the year, month, and day.

```{r}
df_crimes <- mutate(df_crimes, date = as.Date(df_crimes$REPORT_DAT))
```

### Upload Data Frame to CAS

Use the upload.file function to upload df_crimes to the Casuser caslib as a table named crimes.

```{r include=FALSE}
cas.upload.file(s,
                df_crimes,
                casOut=list(name="crimes",
                            caslib="casuser",
                            replace=TRUE),
                importOptions=list(fileType="csv",
                                   encoding="latin1",
                                   guessRows=30000,
                                   vars=list(date=list(informat="yymmdd10.",   
                                                       format="yymmdd10."))))
```

### Aggregate the Data

Use the aggregation.aggregate action to count the number of rows by month and save the resulting output as a table named CrimesByMonth. Create a CAS table object named tbl_month to reference the CrimesByMonth table.

```{r}
cas.aggregation.aggregate(s,                                                                                  
                          table=list(name="crimes",
                                     caslib="casuser",
                                     groupby=list("date")),                                
                          varSpecs=list(list(name="CCN",
                                             subset=list("N"))),
                          ID="date",                                                     
                          interval="MONTH",                                                                  
                          casOut=list(name="CrimesByMonth",                                                
                                      caslib='casuser',
                                      replace=TRUE))

tbl_month <- defCasTable(s, "CrimesByMonth", caslib="casuser")
```

### Convert the CAS Table to an R Data Frame

Convert the CrimesByMonth table to a CAS data frame using the CAS table object tbl_month, and then convert the CAS data frame casdf_month to an R data frame named df_month. Filter df_month to include only months from 2021 and save the result as df_monthFilter. Use head to view the rows.

```{r}
casdf_month <- to.casDataFrame(tbl_month, obs=nrow(tbl_month))
df_month <- data.frame(casdf_month)
df_month
df_monthFilter <- df_month %>% filter(date != 22646)
head(df_monthFilter)
```

### Plot the Data

Use ggplot with the geom_line function to plot the data to show the number of crime incidents by month. Specify df_monthFilter as the first argument. Specify date_f (month) as the X-axis variable and X_CCN_Summary_NObs_ (number of rows) as the Y-axis variable and set the group parameter equal to 1.

```{r}
ggplot(df_monthFilter, aes(x=date_f, y=X_CCN_Summary_NObs_, group=1)) +
  geom_line(size=2, color="lightblue") +
  geom_point(size=3, shape=21, color="lightblue", fill="lightblue", stroke = 2) +
  ylim(1500, 3000) +
  labs(x = "Month", 
       y = "Number of Crime Incidents",
       title = "Number of Crime Incidents by Month",
       subtitle = "Washington DC, 2021") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
```
